---
title: "NYPD Shooting Incidents Analysis"
author: "Data Scince Student"
date: "11/03/2021"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R markdown document for NYPD Shooting Incidents. This document is prepared as an assignment for the course - "Data Science as field" from University of Colorado Boulder.

## Load the NYPD Shooting data from the URL

```{r get_daata}

url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

```

#Read the data from csv:


```{r import_data, message = FALSE}

library(readr)
nypd_cases <- read_csv(url_in)

```

## Remove the extra columns and change the date format:


```{r remove_columns, message = FALSE}

library(tidyverse)
library(lubridate)
nypd_cases <- nypd_cases %>%
  select(-c(X_COORD_CD,Y_COORD_CD,Latitude,Longitude,Lon_Lat,JURISDICTION_CODE)) %>%
  mutate(event_date = mdy(`OCCUR_DATE`))

```
## Analyze the data by the age_group of perpetrator:


```{r age_group, message = FALSE}
nypd_age_group_events <- nypd_cases %>%
  filter(PERP_AGE_GROUP != "UNKNOWN" & PERP_AGE_GROUP != "")

nypd_group_by_age <- nypd_age_group_events %>%
  group_by(PERP_AGE_GROUP) %>%
  summarise(count_incidents = n()) %>%
  ungroup()

nypd_group_by_age <- nypd_group_by_age %>%
  filter(count_incidents > 1)

library(ggplot2)

group_by_age_frame <- data.frame(nypd_group_by_age)

group_by_age_plot<-ggplot(data=group_by_age_frame, aes(x=PERP_AGE_GROUP, y=count_incidents)) +
  geom_bar(stat="identity", fill="steelblue")+
  geom_text(aes(label=count_incidents), vjust=-0.3, size=3.5)+
  theme_minimal()+
  labs(title = "NYPD SHOOTINGS BY AGE GROUP") +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

group_by_age_plot
```

## Analyze the data by race of perpetrator:


```{r race, message = FALSE}

nypd_race_events <- nypd_cases %>%
  filter(PERP_RACE != "UNKNOWN" & PERP_RACE != "")

nypd_group_by_race <- nypd_race_events %>%
  group_by(PERP_RACE) %>%
  summarise(count_incidents = n()) %>%
  ungroup()

nypd_group_by_race <- nypd_group_by_race %>%
  filter(count_incidents > 2)

group_by_race_frame <- data.frame(nypd_group_by_race)

group_by_race_plot<-ggplot(data=group_by_race_frame, aes(x=PERP_RACE, y=count_incidents)) +
  geom_bar(stat="identity", fill="darksalmon")+
  geom_text(aes(label=count_incidents), vjust=-0.3, size=3.5)+
  theme_minimal()+
  labs(title = "NYPD SHOOTINGS BY RACE") +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

group_by_race_plot

```

## Analyze the data by sex of perpetrator:


```{r sex, message = FALSE}

nypd_sex_events <- nypd_cases %>%
  filter(PERP_SEX != "U" & PERP_SEX != "")

nypd_group_by_sex <- nypd_sex_events %>%
  group_by(PERP_SEX) %>%
  summarise(count_incidents = n()) %>%
  ungroup()


nypd_group_by_sex_frame <- data.frame(nypd_group_by_sex)

group_by_sex_plot<-ggplot(data=nypd_group_by_sex_frame, aes(x=PERP_SEX, y=count_incidents)) +
  geom_bar(stat="identity", fill="darkviolet")+
  geom_text(aes(label=count_incidents), vjust=-0.3, size=3.5)+
  theme_minimal()+
  labs(title = "NYPD SHOOTINGS BY SEX") +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

group_by_sex_plot

```

## Model - Statistical murder flag as a function of BORO 

```{r MURDER_FLAG_BORO}

mod_boro <- lm(STATISTICAL_MURDER_FLAG ~ BORO, data = nypd_cases)
summary(mod_boro)
```

## Model - Statistical murder flag as a function of PERP_AGE_GROUP
```{r MURDER_FLAG_PERP_AGE}

mod_age <- lm(STATISTICAL_MURDER_FLAG ~ PERP_AGE_GROUP, data = nypd_age_group_events)
summary(mod_age)
```

We can see from the data that there are lots of values with UNKNOWN and empty values for age groups, race, and sex. I analysed only those values that have valid values. 

The data indicates that the incidents were committed by particular age group, race, and sex. I believe crimes can be controlled by providing necessary livelihood conditions to everyone so that the community is at comparable economic and social level.  

Before looking at the data, I had a personal bias that the shooting crimes were committed by the age group 25-44 but the data proved that the maximum number of incidents were committed by age 18-24.
